---
title: "Two Ways of Specifying Random Intercepts in SPSS's Mixed Model"
author: "Josh Errickson"
output:
  html_document:
    toc: yes
    toc_depth: 4
    toc_float: yes
---

# Introduction

To do: Screenshots of SPSS dialogues.

In SPSS's Mixed Models dialogue, there are two ways to enter random intercepts,
either by the Subjects and Repeated measures dialogue (the first window upon
opening the dialogue) or the Random subdialogue. This document shows how to
generate identical results using either option.

# Loading data in R

We'll use R to get the data, demostrate the model we're trying to fit, and
export the data for SPSS. The data comes from Brady West.

```{r}
#| message: false
library(tidyverse)
ratpup <- readr::read_delim("http://www-personal.umich.edu/~bwest/rat_pup.dat", "\t")
```

Make the treatment binary, and convert both it and sex to numeric.

```{r}
ratpup <- ratpup %>%
  mutate(treat = (treatment != "Control") + 0) %>%
  mutate(female = (sex == "Female") + 0) %>%
  select(-c(treatment, sex)) %>%
  as.data.frame()
```

Write the data into SPSS

```{r}
library(foreign)
write.foreign(ratpup, datafile = "ratpup.sav",
              codefile = "ratpup.sps", package = "SPSS")
```

The mixed model we'll be fitting has a random intercept per litter:

```{r}
#| eval: false
library(lme4)
lmer(weight ~ treat*female + litsize + (1 | litter), data = ratpup, REML = TRUE)
```

# SPSS

Open the file `ratpup.sps` and run the code to load the data.

We'll run two variations using the dialogue Analyze -> Mixed Models -> Linear.

First, ignore the first screen. Place `weight` in Dependent variable; `treat`,
`female` and `litter` into Factors, and `litsize` into Covariates. In the Fixed
subdialogue, enter the main effect for litsize and the main and interaction of
`treat` and `female`. In the Random subdialogue, place `litter` into the Model.
Do **not** check the box "Include intercept".

This dialogue should produce the following syntax:

```
MIXED weight BY female litter treat WITH litsize
  /CRITERIA=CIN(95) MXITER(100) MXSTEP(10) SCORING(1) SINGULAR(0.000000000001)
  HCONVERGE(0, ABSOLUTE) LCONVERGE(0, ABSOLUTE) PCONVERGE(0.000001, ABSOLUTE)
  /FIXED=female treat female*treat litsize | SSTYPE(3)
  /METHOD=REML
  /PRINT=SOLUTION
  /RANDOM=litter | COVTYPE(VC).
```


In this model, `litter` was chosen as "subject" in the first screen. Random
Intercept was included but not `litter`, and `litter` was included in "Subject
groupings"


Next, we'll make use of the first screen. In the first screen, place `little`
into Subjects. Enter the variables as described above.^[`litter` does not have
to be placed in Factors anymore, but it will not affect anything if it is.] In
the Random subdialogue, check "Include intercept". Do **not** enter `litter`
into Model, but at the bottom under Subject Groups, enter it into Combinations.

The syntax is:

```
MIXED weight BY female treat WITH litsize
  /CRITERIA=CIN(95) MXITER(100) MXSTEP(10) SCORING(1) SINGULAR(0.000000000001)
  HCONVERGE(0, ABSOLUTE) LCONVERGE(0, ABSOLUTE) PCONVERGE(0.000001, ABSOLUTE)
  /FIXED=female treat female*treat litsize | SSTYPE(3)
  /METHOD=REML
  /PRINT=SOLUTION
  /RANDOM=INTERCEPT | SUBJECT(litter) COVTYPE(VC).
```

The difference is

```
  /RANDOM=litter | COVTYPE(VC).
```

versus

```
  /RANDOM=INTERCEPT | SUBJECT(litter) COVTYPE(VC).
```

but both models are identical.
