---
title: "Log Transform Interpretation"
author: "Josh Errickson"
output:
  html_document:
    toc: yes
    toc_depth: 4
    toc_float: yes
---

This document discusses interpreting OLS regression coefficients when the response and/or the predictor is log transformed.

## Data

```{r}
#install.packages("ISLR")
library(ISLR)
data(Hitters)
set.seed(1) # random seed for reproducibility
```

Dataset is the `Hitters` dataset from the `ISLR` package, which contains batting statistics for players in the '86 and '87 seasons.

The two variables we'll be log transforming are `Salary` and `Hits`.

## Log transforming the response

```{r}
summary(mod1 <- lm(log(Salary) ~ AtBat + Hits + HmRun + Years + League, data = Hitters))
```

### Binary predictor

Let's create two new observations with the same value of all covariates, but different values of the binary predictor `League`.
```{r}
(newH <- data.frame(AtBat = sample(1:50, 1),
                    Hits = sample(1:30, 1),
                    HmRun = sample(1:10, 1),
                    Years = sample(1:10, 1),
                    League = c("A", "N")))
```

Get the predicted values, and exponentiate them to put them on the scale of `Salary` rather than `log(Salary)`.

```{r}
(p <- exp(predict(mod1, newdata = newH)))
```

```{r}
exp(mod1$coef["LeagueN"])
p[1]*exp(mod1$coef["LeagueN"])
```

Therefore the interpretation is

> The predicted average salary is `r round(100*(exp(mod1$coef["LeagueN"]) - 1), 2)`\% higher for a National League hitter than an American League hitter (with other predictors held constant)."

### Continuous predictor

Generate new observations with one more `HmRun`, and obtain predicted values on `Salary` scale.

```{r}
(newH <- data.frame(AtBat = sample(1:50, 1),
                    Hits = sample(1:30, 1),
                    HmRun = c(40,41),
                    Years = sample(1:10, 1),
                    League = "A"))
(p <- exp(predict(mod1, newdata = newH)))
```

```{r}
exp(mod1$coef["HmRun"])
p[1]*exp(mod1$coef["HmRun"])
```

The interpretation is similar to the binary case,

> The predicted average salary will increase by `r round(100*(exp(mod1$coef["HmRun"]) - 1), 2)`\%  for each additional home run a hitter has (with other predictors held constant)."


## Log transforming a predictor

Obviously we don't need to be concerned with the binary predictor here.
```{r}
summary(mod2 <- lm(Salary ~ AtBat + log(Hits) + HmRun + Years + League, data = Hitters))
```

Generate the two hitters, with one having a 10\% higher number of hits than the other.
```{r}
(newH <- data.frame(AtBat = sample(1:50, 1),
                    Hits = c(14, 14*1.1),
                    HmRun = sample(1:10, 1),
                    Years = sample(1:10, 1),
                    League = "A"))
(p <- predict(mod2, newdata = newH))
log(1.1)*mod2$coef["log(Hits)"]
p[1] + log(1.1)*mod2$coef["log(Hits)"]
```
Interpreted as

> The predicted average salary decreases by $`r abs(round(log(1.1)*mod2$coef["log(Hits)"], 2))` when the number of hits increases by 10\% (when all other predictors are constant).

## Log transforming both


```{r}
summary(mod3 <- lm(log(Salary) ~ AtBat + log(Hits) + HmRun + Years + League, data = Hitters))
```

Obviously, unlogged coefficients (such as `League` or `HmRun`) have the same interpretation as before, though with the assumption of the new model. (`mod3` is different from `mod1` because the relationship is linear with `Hits` on different scales.)

Generate the two data points with a 20\% higher number of hits for one.
```{r}
(newH <- data.frame(AtBat = sample(30:50, 1),
                    Hits = c(15, 15*1.2),
                    HmRun = sample(1:10, 1),
                    Years = sample(1:10, 1),
                    League = "A"))
```

Exponentiate the predicted values to get back on the scale of `Salary`.

```{r}
(p <- exp(predict(mod3, newdata = newH)))
1.2^mod3$coef["log(Hits)"]
p[1]*(1.2^mod3$coef["log(Hits)"])
```

> The predicted average salary will decrease by `r round(100*(1 - 1.2^mod3$coef["log(Hits)"]), 2)`\% when a hitter's number of hits increases by 20\% (with other predictors held constant).

The  `r round(100*(1 - 1.2^mod3$coef["log(Hits)"]), 2)` value comes from 1.0 -  `r round( 1.2^mod3$coef["log(Hits)"], 3)`.

## Summary

With a log-transformed response, a linear change in the predictor leads to a percentage change in the response.

With a log-transformed predictor, a percentage change in the predictor leads to a linear change in the response.

With both a log-transformed response and a log-transformed predictor, a percentage change in the predictor leads to a percentage change in the response.
