---
title: "Response Surface Plot"
author: "Josh Errickson"
output:
  html_document:
    toc: yes
    toc_depth: 4
    toc_float: yes
---

# Introduction

Response surface analysis is just regression with an interaction. Typically the model fit is

\[
    y = \beta_0 + \beta_1x_1 + \beta_2x_2 + \beta_3x_1x_2 + \beta_4x_1^2 + \beta_5x_2^2,
\]

but in general any interaction should work for the below visualization.

# Using `rsm::persp.lm`

The "rsm" package allegedly does response surface modeling automatically, but we
can use it just for the `rsm::persp.lm` function (which can just be referred to
as `persp` after loading the package). Note that `stats::persp` does *not* work
directly with a model object; you'd need to do a few extra steps to generate a
matrix first.

```{r, message = FALSE}
library(rsm)

data(mtcars)

mod1 <- lm(mpg ~ disp + qsec + disp:qsec + I(disp^2) + I(qsec^2), data = mtcars)

rsm::persp.lm(mod1, disp ~ qsec, zlab = "mpg", theta = 120, phi = 10, r = 10)
```

(Note: `persp.lm` does not support other types of models such as mixed models
from `lmer`, however plotly below should support all types of models that
[`emmeans` supports](https://cran.r-project.org/web/packages/emmeans/vignettes/models.html).)

# Using plotly

Alternately, [plotly](https://plotly.com/r/) can be used to create an
interactive plot. The data needs to be set up in a bit of an odd fashion,
instead of a matrix of x-y-z triads, a list is created with the unique x values,
the unique y values, and a matrix of predicted values of the outcome containing
all pairwise values of x and y. We use `emmeans` to generate these predicted
values.

```{r, message = FALSE}
library(plotly)
library(emmeans)


# Set up grid
q <- 14:23
d <- (1:10)*50

em <- as.data.frame(emmeans(mod1, ~ qsec*disp, at = list(qsec = q, disp = d)))

# x & y are unique values, z is len(x) x len(y) matrix
dd <- list(q = q,
           d = d,
           m = matrix(em$emmean,
                      nrow = length(q),
                      ncol = length(d),
                      byrow = TRUE))

plot_ly(x = dd$q, y = dd$d, z = dd$m, type = "surface") |>
  layout(scene = list(xaxis=list(title = "qsec"),
                      yaxis=list(title = "disp"),
                      zaxis=list(title = "mpg")))
```


# A Warning about Interpretation

Response surface plots can be extremely misleading due to extrapolation.
Consider the plots above. One thing we would be tempted to interpret from the
plot is that the lowest mileage cars (~5-10 MPG) occurs when `disp` and `qsec`
are at their maximum. However, this is complete extrapolation. We can look at
the predicted values for the cars in the data:

```{r}
summary(predict(mod1))
```

The lowest predicted mileage is `r round(min(predict(mod1)), 2)`, yet the
response surface dips below 5! What's happening is that there are no cars in the
car with both high `disp` and high `qsec`, so that area of the response curve is
entirely extrapolated - we have no idea what should be occurring there.

```{r}
plot(mtcars$disp ~ mtcars$qsec)
text(21, 400, "There's no data in this region!", col = "red")
```

Finally, let's add the predicted values onto the response surface to further
visualize the areas of the response surface that is pure extrapolation. (Note
that I add one to the predicted values (`predict(m) + 1`) to move the points
slightly above the surface, improving visualization.)


```{r}
df <- data.frame(qsec = mtcars$qsec,
                 disp = mtcars$disp,
                 predict = predict(mod1) + 1)


plot_ly(x = dd$q, y = dd$d, z = dd$m, type = "surface") |>
  layout(scene = list(xaxis=list(title = "qsec"),
                      yaxis=list(title = "disp"),
                      zaxis=list(title = "mpg"))) |>
  add_trace(x = df$qsec, y = df$disp, z = df$predict, mode = "markers",
            type = "scatter3d",
            marker = list(size = 10, color = "red", symbol = 104))
```
