---
title: "Visualizing the issue with Collinear Predictors"
author: "Josh Errickson"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
```

# Introduction

While there are many resources out there to describe the issues arising with multicollinearity in independent variables, there's a visualization for
the problem that I came across in a book once and haven't found replicated online. This replicates the visualization.

# Set-up

Let \(Y\) be a response and \(X_1\) and \(X_2\) be the predictors, such that

\[
  Y_i = \beta_0 + \beta_1X_{1i} + \beta_2X_{2i} + \epsilon_i
\]

for individual \(i\).

For simplicity, let's say that

\[
\beta_0 = 0,\\
\beta_1 = 1,\\
\beta_2 = 1.
\]

I carry out a simulation by generating 1,000 simulations with a given correlation between predictors and obtain their coefficients.

```{r, eval = FALSE}
reps <- 1000
n <- 100
save <- matrix(nrow = reps, ncol = 3)

for (i in 1:reps) {
  x1 <- rnorm(n)
  x2 <- rnorm(n)
  y <- x1 + x2 + rnorm(n)

  mod <- lm(y ~ x1 + x2)
  save[i,] <- c(coef(mod)[-1], cor(x1, x2))
}
```

The line `x2 <- rnorm(n)` gets replaced with `x2 <- x1 + rnorm(n, sd = _)`, where the `_` is replaced with difference values to induce more correlation between `x1` and `x2`.

Following these simulations, the coefficients are plotted against each out.

# Simulation {.tabset}

## No Collinearity

```{r, echo = FALSE}
reps <- 1000
n <- 100
save <- matrix(nrow = reps, ncol = 3)

for (i in 1:reps) {
  x1 <- rnorm(n)
  x2 <- rnorm(n)
  y <- x1 + x2 + rnorm(n)

  mod <- lm(y ~ x1 + x2)
  save[i,] <- c(coef(mod)[-1], cor(x1, x2))
}
save <- data.frame(save)
names(save) <- c("b_1", "b_2")
```

The average correlation between \(X_1\) and \(X_2\) is `r round(mean(save[,3]), 3)`.

```{r, echo = FALSE, fig.height = 7, fig.width = 7}
ggplot(save, aes(x = b_1, y = b_2)) +
  geom_point() +
  geom_point(aes(x = 1, y = 1), col = "red", size = 3) +
  ylab(bquote(hat(beta)[2])) +
  xlab(bquote(hat(beta)[1]))
```

The red dot represents the true coefficients. We see no relationship between the estimated coefficients, and each are well centered around the truth.

## Moderate Collinearity of .5

```{r, echo = FALSE}
reps <- 1000
n <- 100
save <- matrix(nrow = reps, ncol = 3)

for (i in 1:reps) {
  x1 <- rnorm(n)
  x2 <- x1 + rnorm(n, sd = 1.5)
  y <- x1 + x2 + rnorm(n)

  mod <- lm(y ~ x1 + x2)
  save[i,] <- c(coef(mod)[-1], cor(x1, x2))
}
save <- data.frame(save)
names(save) <- c("b_1", "b_2")
```

The average correlation between \(X_1\) and \(X_2\) is `r round(mean(save[,3]), 3)`.

```{r, echo = FALSE, fig.height = 7, fig.width = 7}
ggplot(save, aes(x = b_1, y = b_2)) +
  geom_point() +
  geom_point(aes(x = 1, y = 1), col = "red", size = 3) +
  ylab(bquote(hat(beta)[2])) +
  xlab(bquote(hat(beta)[1]))
```

While there is some semblance of a relationship, it is not very strong.

## High Collinearity of .9

```{r, echo = FALSE}
reps <- 1000
n <- 100
save <- matrix(nrow = reps, ncol = 3)

for (i in 1:reps) {
  x1 <- rnorm(n)
  x2 <- x1 + rnorm(n, sd = .5)
  y <- x1 + x2 + rnorm(n)

  mod <- lm(y ~ x1 + x2)
  save[i,] <- c(coef(mod)[-1], cor(x1, x2))
}
save <- data.frame(save)
names(save) <- c("b_1", "b_2")
```

The average correlation between \(X_1\) and \(X_2\) is `r round(mean(save[,3]), 3)`.

```{r, echo = FALSE, fig.height = 7, fig.width = 7}
ggplot(save, aes(x = b_1, y = b_2)) +
  geom_point() +
  geom_point(aes(x = 1, y = 1), col = "red", size = 3) +
  ylab(bquote(hat(beta)[2])) +
  xlab(bquote(hat(beta)[1]))
```

There's a strong relationship now, such that if one coefficient is incorrectly high, the other is incorrectly low.

## Extremely High Collinearity of .99

```{r, echo = FALSE}
reps <- 1000
n <- 100
save <- matrix(nrow = reps, ncol = 3)

for (i in 1:reps) {
  x1 <- rnorm(n)
  x2 <- x1 + rnorm(n, sd = .1)
  y <- x1 + x2 + rnorm(n)

  mod <- lm(y ~ x1 + x2)
  save[i,] <- c(coef(mod)[-1], cor(x1, x2))
}
save <- data.frame(save)
names(save) <- c("b_1", "b_2")
```

The average correlation between \(X_1\) and \(X_2\) is `r round(mean(save[,3]), 3)`.

```{r, echo = FALSE, fig.height = 7, fig.width = 7}
ggplot(save, aes(x = b_1, y = b_2)) +
  geom_point() +
  geom_point(aes(x = 1, y = 1), col = "red", size = 3) +
  geom_abline(intercept = 2, slope = -1, color = "red") +
  ylab(bquote(hat(beta)[2])) +
  xlab(bquote(hat(beta)[1]))
```

We now see a very strong relationship between the coefficients. Think of it this way: With such high correlation, essentially \(X_1 = X_2\), which is
represented by the red line (slope of -1). We can approximate our model:

\begin{aligned}
  Y_i &= \beta_0 + \beta_1X_{1i} + \beta_2X_{2i} + \epsilon_i\\
  &\approx \beta_0 + (\beta_1 + \beta_2)X_{1i} + \epsilon_i\\
  &\approx \beta_0 + (\beta_1 + \beta_2)X_{2i} + \epsilon_i
\end{aligned}

In other words, the model has that \(\beta_1 + \beta_2 = 2\). The red line in the above plot is all values which agree with that (e.g., \(\beta_1 =
1\) and \(\beta_2 = 1\) or \(\beta_1 = 3\) and \(\beta_2 = -1\)). So while all of those models would have the same predictive power for \(Y\), they
would have drastically different interpretations depending on where along that red line they fall.
